SFJ: An implementation of Semantic Featherweight Java

### 2020-04-27 conference/camera ready/background.tex
first pass on paper after revision

Following FJ \cite{featherweight}, we rule out ill-formed programs, such as declaring a constructor named $B$ within a class named $A$; or multiple fields or methods having the same name; or fields having the same type as the type of the class they are defined in.

>>>>

Following FJ \cite{featherweight}, we rule out ill-formed programs, such as declaring a constructor named $B$ within a class named $A$, or multiple fields or methods having the same name, or fields having the same type as the class they are defined in.

### 2020-04-27 conference/camera ready/paper.tex
first pass on paper after revision

%% For section renaming in autoref
\addto\extrasenglish{%
  \renewcommand{\sectionautorefname}{\S}%
  \renewcommand{\subsectionautorefname}{\S}%
  \renewcommand{\algorithmautorefname}{Algorithm}%
  %\renewcommand{\exampleautorefname}{Ex.}%
  \renewcommand{\equationautorefname}{Mapping}%
}

@@

\section{The SFJ Language}
\label{sec:design}
\subsection{On Valid Type Definitions}
\label{sec:valid}
Since we want to use types $\tau$ in practice in SFJ, we restrict them to \emph{finite trees} whose leaves are basic types $\btypes$ \autoref{sec:terms} with no cycles.
For example, a recursive type $\alpha = [a : \alpha]$ denotes an infinite program tree \textbf{new} $C$(\textbf{new} $C$($\cdots{}$)), hence we rule it out as it is not inhabitable.
Similarly, the types $\alpha = [b: \beta]$, $\beta = [a = \alpha]$ create a cycle and thus would not be inhabitable.
Notice that these types can be defined and inhabited in Java by assigning \emph{null} to all fields in a class, however they are not useful in practice.

SFJ is implemented using ANTLR \cite{parr2013}.
We start by defining the grammar of the language in Extended Backus-Naur Form (EBNF), following \autoref{sec:types} and by running ANTLR, we can automatically generate a parser for SFJ and extend it in order to implement the required checks for our types and type system.
Running the parser on an SFJ program returns an abstract syntax tree (AST) of that program.

When visiting the AST, we check if the program is well-formed, following the intuition at the end of \autoref{sec:background}.
We mark any classes containing fields typed with only basic types as \emph{resolved} otherwise, as \emph{unresolved}.
Using this information, \autoref{alg:types} checks if the type definitions in a program are valid, namely, if they are finite trees whose leaves are basic types with no cycles.
%
At each iteration of the algorithm we can resolve at least one type, until all the types in the SFJ program are resolved, and return $True$.
However, if there is at least one type unresolved, meaning there is a cycle in the type definition, we return $False$.

@@

    \DontPrintSemicolon
    \Input{$classes$, the set of classes marked $resolved$ if their fields contain only basic types, $unresolved$ otherwise.}
    \Output{$True$ if all classes are valid type definitions, $False$ otherwise.}
    \BlankLine
    \Begin{
        \Do{$resolutionOccured = True$}{ %\tcc {comment}
            $resolutionOccured \longleftarrow False$ \;
            \For{class that is unresolved in classes}{
                $resolved \longleftarrow True$ \;
                \For{field in class that contains a class type}{
                    \If{type of field is unresolved}{
                        $resolved \longleftarrow False$
                    }

@@

    \DontPrintSemicolon
    \Input{$classes$, the set of classes in an SFJ program for which we have not yet defined the subtyping relation.\;
    $relation$, the mapping of types to the set of subtypes, initially being Mapping \ref{eq:initialSubtype}. \;
    }
    \Begin{
        \Fn{generateRelation(classes: List<Class>)}{
            $unprocessed: List<Class>\ \longleftarrow []$

@@

It is worth noticing that the subtyping algorithm finds all nominal and structural subtypes of a given type. This is due to the fact that all pairs of types are inspected.
Recall from \autoref{sec:intro} that nominal subtyping is name-based and given by the class hierarchy defined by the programmer, whether structural subtyping is structure-based and given by the set-inclusion of fields and methods. In particular, it is contra-variant with respect to this set-inclusion.
\autoref{alg:subtyping} finds all structural subtypes of a given class because it checks that its fields and methods are a superset of existing types in $relation$. For example, all classes are structural subtypes of type $empty = []$.
On the other hand, it also finds all nominal subtypes because a class inherits all fields and methods of its superclass and as such its fields and methods are a superset of its superclass.
This means that checking for structural subtyping is enough because nominal subtyping will be captured due to inheritance of fields and methods.

Finally, a note on complexity.
The complexity of \autoref{alg:types} is $\mathcal{O}(n)$, and the complexity of \autoref{alg:subtyping} is $\mathcal{O}(n^{2})$, with $n$ being the size of the input.
The reason for a quadratic complexity of \autoref{alg:subtyping} is due to the symmetric check of structural subtyping between a $class$ and an $existingClass$ in $relation$.
Notice that if we were to only work with nominal subtyping, then we only require traversing the class hierarchy once,  which gives an $\mathcal{O}(n)$ complexity.

@@

On the other hand, while nominal subtyping can avoid the above problem, it can introduce others and in particular, a developer can define an overridden method to perform the opposite logic to what the super class is expecting, as illustrated by the following classes in Java:
\begin{equation*}
    \begin{array}{ll}
        \syndecl{A}{\mathit{Object}} {
         &
            \qquad
            \syndecl{B}{A} {
                \\
               \qquad \qquad\ldots \qquad
         &
                \qquad\qquad\qquad \ldots
                \\
                \quad \textbf{int}\ n;\qquad
         &
                \qquad\quad \textbf{int}\ length()\{\ \synret{-n};\ \}
                \\
                \quad \textbf{int}\ length()\{\ \synret{n};\ \}\qquad
         &
                \qquad}
                \\   }
    \end{array}
\end{equation*}
Both approaches have their pros and cons, and they leave an expectation on the developer to use the logic behind subtyping correctly when writing code. Hence, the integration of both approaches in SFJ makes it possible to  overcome these drawbacks, as one can choose on which subtyping to focus for a given task.

@@

In order to type check an argument passed to the (outside) $diagonal$ method, at compile time we build a collection of types $\{type_{1},\ type_{2},\ldots\}$ which are class types where the (inside) $diagonal$ method is defined.
As such, we iterate over the list of \emph{classes} in an SFJ program (as we did in \autoref{alg:subtyping})) to check for the required method.
The resulting collection of types is the union of all classes where $diagonal$ is defined together with their subtypes $([\![type_{1}]\!] \cup [\![type_{2}]\!] \cup \ldots)$, where each $[\![type_{i}]\!]$ denotes a mapping of $type_i$ to the set of its subtypes, similar to \autoref{eq:sugtypePolygons}.

@@

We can therefore use method types to statically include or exclude a portion of our class hierarchy.
However, unlike the use of interfaces, as in one of the proposed Java approaches in \autoref{polygons}, the values that can be accepted by a method type do not have to be related to each other in any way in the class hierarchy.
This indeed is useful if we are dealing with legacy code as we can still accept all classes where \emph{diagonal} is defined, without having to go back and add interface implementations.

@@

\subsection{Code Generation}
\label{sec:code_generation}
%%
SFJ only includes the typechecking component of the language. We provide a ``proof'' of the code generation algorithm, which is work in progress at this stage.

Given the similarity of SFJ to Java, our approach is to translate an SFJ program into Java bytecode and then run it on the Java Virtual Machine (JVM) \cite{lindholm2013}. This is a standard approach used also by other object-oriented languages, for example, Kotlin\footnote{https://kotlinlang.org/}.

The main challenge in translating SFJ into bytecode is translating types using boolean connectives.

For example, a field $f_1$ of type \textbf{int or bool}, will be translated as two Java fields, one of type \textbf{int} and one of type \textbf{bool}, and only one of the two types will be inhabited by a value. 
%
In order to achieve this, we first analyse our program and reduce the boolean types by keeping only the alternatives which actually get used in the program.
For example, if the field of type \textbf{int or bool} only ever gets initialised with a boolean value, we can reduce it and make it a field of a single type \textbf{bool}.
%Similarly, basic types can be reduced to their largest type without consequence, such as \textbf{int and not bool} can be reduced to just \textbf{int} as we have already type checked all literal values passed to such a type.
%
After this reduction phase, we then consider the remaining types which use boolean connectives and could not be reduced.
On the example above, consider again field $f_1$ of type \textbf{int or bool}. This will be translated as two fields \textbf{int}$\_{f_1}$ and \textbf{bool}$\_{f_1}$ with the corresponding types.
In order to initialise these fields, we use the constructor overloading capabilities of the JVM to generate an overloaded version for each alternative type.
In each constructor, only the field that matches the type of the parameter is initialised with all other fields set to $null$.
%
To access the field of an object, we generate code that checks each alternative of the field if it is non-null and includes the rest of the code generation for the expression for each branch. At run time, only one branch will be true, and this is the branch of generated code which will be executed.

For methods, we also use the overloading capabilities of the JVM to define an  overloaded method for each type in the expanded method parameter, all with the same method body.
Depending on which alternative the argument inhabits at run time, a different method will be dispatched to.
%
%For structural subtyping, we create an new nominal subtype with the same implementation for each structural subtype a type has, so that it is accepted as a proper subtype in the JVM.
Like methods with boolean types, we similarly implement methods with method types, as we discussed in \autoref{sec:methods}, by defining an overloaded alternative for each type that implements the specified method.

This concludes the code generation phase for all expressions in SFJ.

@@

On the implementation side, in addition to $\mathbb{C}$Duce, to the best of our knowledge there are few work who have developed tools.
%
Muehlboeck and Tate \cite{Muehlboeck2018} define a syntactic framework with boolean connectives which has been implemented in the Ceylon programming language \cite{Ceylon2016}.
%
Ancona and Corradi \cite{AnconaC16} define semantic subtyping for an imperative object-oriented language with mutable fields. In our framework we are considering only the functional fragment of Java, which is FJ, and as a result the semantic subtyping framework is simpler. The authors also propose a prototype implementation of their subtyping algorithm.
%
Chaudhuri \emph{et al.} \cite{ChaudhuriVGRL17} present the design and implementation of \textsc{Flow}, which is a type checker for JavaScript. They use boolean connectives \textbf{and}, \textbf{not}, \textbf{or} for their predicates, however they do not define semantic subtyping for their language.

>>>>

%% For section renaming in autoref
\addto\extrasenglish{%
  \renewcommand{\sectionautorefname}{\S}%
  \renewcommand{\subsectionautorefname}{\S}%
  \renewcommand{\algorithmautorefname}{Algo.}%
  %\renewcommand{\exampleautorefname}{Ex.}%
  \renewcommand{\equationautorefname}{Map}%
}

@@

\section{The SFJ Language}
\label{sec:design}
\subsection{On Valid Type Definitions}
\label{sec:valid}
Since we want to use types $\tau$ to program in SFJ, we restrict them to \emph{finite trees} whose leaves are basic types $\btypes$ \autoref{sec:terms} with no cycles.
For example, a recursive type $\alpha = [a : \alpha]$ denotes an infinite program tree \textbf{new} $C$(\textbf{new} $C$($\cdots{}$)), hence we rule it out as it is not inhabitable.
Similarly, the types $\alpha = [b: \beta]$, $\beta = [a = \alpha]$ create a cycle and thus would not be inhabitable.
Notice that these types can be defined and inhabited in Java by assigning \emph{null} to all fields in a class, however they are not useful in practice.

SFJ is implemented using ANTLR \cite{parr2013}.
We start by defining the grammar of the language in Extended Backus-Naur Form (EBNF), following \autoref{sec:types} and by running ANTLR, we can automatically generate a parser for SFJ and extend it in order to implement the required checks for our types and type system.

Running the parser on an SFJ program returns an abstract syntax tree (AST) of that program.
When visiting the AST, we check if the program is well-formed, following the intuition at the end of \autoref{sec:background}.
We mark any classes containing only fields typed with basic types as \emph{resolved}, otherwise we mark them as \emph{unresolved}.
Using this information, \autoref{alg:types} checks if the type definitions in a program are valid, namely if they are finite trees whose leaves are basic types with no cycles.
%
The algorithm returns $True$ only if all the types in the program are resolved, otherwise it returns $False$, meaning there is at least one type definition which is invalid and contains a cycle.

@@

    \DontPrintSemicolon
    \Input{$classes$, the set of classes in an SFJ program marked $resolved$, if their fields contain only basic types, $unresolved$ otherwise.}
    \Output{$True$ if all classes are valid type definitions, $False$ otherwise.}
    \BlankLine
    \Begin{
        \Do{$resolutionOccured = True$}{ %\tcc {comment}
            $resolutionOccured \longleftarrow False$ \;
            \For{class that is unresolved in classes}{
                $resolved \longleftarrow True$ \;
                \For{field in class that contains a class type}{
                    \If{type of field is unresolved}{
                        $resolved \longleftarrow False$
                    }

@@

    \DontPrintSemicolon
    \Input{$classes$, the set of classes in an SFJ program.\\
    $relation$, the mapping of types to the set of subtypes, initially Map \ref{eq:initialSubtype}. \;
    }
    \Begin{
        \Fn{generateRelation(classes: List<Class>)}{
            $unprocessed: List<Class>\ \longleftarrow []$

@@

It is worth noticing that the subtyping algorithm finds all nominal and structural subtypes of a given type. This is due to the fact that all pairs of types are inspected.
Recall from \autoref{sec:intro} that nominal subtyping is name-based and given by the class hierarchy defined by the programmer, whereas structural subtyping is structure-based and given by the set-inclusion of fields and methods. In particular, structural subtyping is contra-variant with respect to this set-inclusion.
\autoref{alg:subtyping} finds all structural subtypes of a given class because it checks that its fields and methods are a superset of existing types in $relation$. For example, all classes are structural subtypes of type $empty = []$.
On the other hand, it also finds all nominal subtypes because a class inherits all fields and methods of its superclass and as such its fields and methods are a superset of its superclass.
This means that checking for structural subtyping is enough because nominal subtyping will be captured due to inheritance of fields and methods.

Finally, a note on complexity.
The complexity of \autoref{alg:types} is $\mathcal{O}(n)$, and the complexity of \autoref{alg:subtyping} is $\mathcal{O}(n^{2})$, with $n$ being the size of the input.
The reason for a quadratic complexity of \autoref{alg:subtyping} is due to the symmetric check of structural subtyping between a $class$ and an $existingClass$ in $relation$.
Notice that if we were to only work with nominal subtyping, then we would only require traversing the class hierarchy once,  which gives an $\mathcal{O}(n)$ complexity.

@@

On the other hand, while nominal subtyping can avoid the above problem, it can introduce others and in particular, a developer can define an overridden method to perform the opposite logic to what the super class is expecting, as illustrated by the following classes in Java:
\begin{equation*}
    \begin{array}{ll}
        \syndecl{A}{\mathit{Object}} {
         &
            \qquad
            \syndecl{B}{A} {
                \\
               \qquad \qquad\ldots \qquad
         &
                \qquad\qquad\qquad \ldots
                \\
                \quad \textbf{int}\ n;\qquad
         &
                \qquad\quad \textbf{int}\ length()\{\ \synret{-n};\ \}
                \\
                \quad \textbf{int}\ length()\{\ \synret{n};\ \}\qquad
         &
                \qquad}
                \\   }
    \end{array}
\end{equation*}
Both approaches have their pros and cons, and they leave an expectation on the developer to use the logic behind subtyping correctly when writing code.
The integration of both subtyping approaches in SFJ gives the developer the freedom to choose the most suitable subtyping relation to use for a given task.

@@

In order to type check an argument passed to the (outside) $diagonal$ method, at compile time we build a collection of types $\{type_{1},\ type_{2},\ldots\}$ which are class types where the (inside) $diagonal$ method is defined.
As such, we iterate over the list of \emph{classes} in an SFJ program (as we did in \autoref{alg:subtyping}) to check for the required method.
The resulting collection of types is the union of all classes where $diagonal$ is defined together with their subtypes $([\![type_{1}]\!] \cup [\![type_{2}]\!] \cup \ldots)$, where each $[\![type_{i}]\!]$ denotes a mapping of $type_i$ to the set of its subtypes, similar to \autoref{eq:sugtypePolygons}.

@@

We can therefore use method types to statically include or exclude a portion of our class hierarchy.
However, unlike with interfaces as in \autoref{polygons}, the values that can be accepted by a method type do not have to be related to each other in any way in the class hierarchy.
This indeed is useful if we are dealing with legacy code as we can still accept all classes where \emph{diagonal} is defined, without having to go back and add interface implementations.

@@

\subsection{Code Generation}
\label{sec:code_generation}
%%
SFJ only includes the typechecking component of the language. In this section, we provide a sketch of the code generation algorithm, which is work in progress.

Given the similarity of SFJ to Java, our approach is to translate an SFJ program into Java bytecode and then run it on the Java Virtual Machine (JVM) \cite{lindholm2013}. This is a standard approach also used by other object-oriented languages, for example, Kotlin\footnote{https://kotlinlang.org/}.

The main challenge in translating SFJ into bytecode is translating types using boolean connectives.
%
For example, a field $f_1$ of type \textbf{int or bool}, will be translated as two Java fields, one of type \textbf{int} and one of type \textbf{bool}, and only one of the two types will be inhabited by a value. 
%
In order to achieve this, we first analyse our program and reduce the boolean types by keeping only the alternatives which actually get used in the program.
For example, if the field of type \textbf{int or bool} only ever gets initialised with a boolean value, we can reduce it and make it a field of a single type \textbf{bool}.
%Similarly, basic types can be reduced to their largest type without consequence, such as \textbf{int and not bool} can be reduced to just \textbf{int} as we have already type checked all literal values passed to such a type.
%
After this reduction phase, we then consider the remaining types which use boolean connectives and could not be reduced.
On the example above, consider again field $f_1$ of type \textbf{int or bool}. This will be translated as two fields \textbf{int}$\_{f_1}$ and \textbf{bool}$\_{f_1}$ with the corresponding types.
In order to initialise these fields, we use the constructor overloading capabilities of the JVM to generate an overloaded version for each alternative type.
In each constructor, only the field that matches the type of the parameter is initialised with all other fields set to $null$.
%
To access the field of an object, we generate code that checks each alternative of the field if it is non-null and includes the rest of the code generation for the expression for each branch. At run time, only one branch will be true, and this is the branch of generated code which will be executed.

For methods, we also use the overloading capabilities of the JVM to define an  overloaded method for each type in the expanded method parameter, all with the same method body.
Depending on which alternative the argument inhabits at run time, a different method will be dispatched to.
%
%For structural subtyping, we create an new nominal subtype with the same implementation for each structural subtype a type has, so that it is accepted as a proper subtype in the JVM.
Like methods with boolean types, we similarly implement methods with method types, as we discussed in \autoref{sec:methods}, by defining an overloaded alternative for each type that implements the specified method.
%
This concludes the code generation phase for all expressions in SFJ given in \autoref{sec:terms}.

@@

Regarding implementations of semantic subtyping, to the best of our knowledge, there are only a few works in the literature.
%
Muehlboeck and Tate \cite{Muehlboeck2018} define a syntactic framework with boolean connectives which has been implemented in the Ceylon programming language \cite{Ceylon2016}.
%
Ancona and Corradi \cite{AnconaC16} define semantic subtyping for an imperative object-oriented language with mutable fields. In our framework we are considering only the functional fragment of Java, which is FJ, and as a result the semantic subtyping framework is simpler. The authors also propose a prototype implementation of their subtyping algorithm.
%
Chaudhuri \emph{et al.} \cite{ChaudhuriVGRL17} present the design and implementation of \textsc{Flow}, which is a type checker for JavaScript. They use boolean connectives \textbf{and}, \textbf{not}, \textbf{or} for their predicates, however they do not define semantic subtyping for their language.
